\section{A Strong Saxon Philippic Against Luther}

“You have entitled your work against the diet, “Warning to my
dear Germans.” It would have been more correct if you had entitled
it, Seduction and Incitement to Rebellion and Insurrection. For you
are plainly bent upon inciting the people to revolt against the Emperor
and all in authority. You proceed with artful snares, with lies
and outcries, with a veritable specter of assertions. The devil is called
upon to assist you on every page. May those pass judgment on
this pamphlet who call you a living saint and say you are possessed
of the spirit of God.” Thus far Cochlaeus’ quotation from Duke
George.

The following quotations are excerpted from the other works mentioned
above:

Your “Glosses” contain as many lies as words. You are an infamous apostate
and pervert the truth when you say that the married state is prohibited
in the Catholic Church. You lie when you maintain that capital punishment
is contemplated for those who do not use consecrated salt. You lie when you
assert that Duke George of Saxony and the Elector of Brandenburg
promised to supply the Emperor with 5000 horses against the Lutherans. You
lie when you aver that the Emperor is resolved to make war upon
you Lutherans for the purpose of suppressing the gospel. You lie like a conscious,
faithless, and perjured scoundrel, when you state that the Catholic princes
have formed a league at Dessau for the purpose of attacking you. Thus you
heap lie upon lie. Our princes formed a league for the sake of preserving the
peace, only to further the welfare of their people, and to the disadvantage
of nobody, as events have shown. You inveigh against all papists, including
the Emperor, traducing them as traitors, miscreants, sacrilegious
blood-hounds. At the same time you falsely pretend not to have spoken
against the Emperor or any other ruler; it is as if one said that he intended
to produce white, but in reality produced black.

The noble-minded Emperor’s sole intention, formed after due deliberation,
is to make order prevail in the Church. He will not allow priests to consort
with dissolute women. You, evil-minded apostate, interpret this intention in
a way which the devil in hell could not surpass in malice.

Everyone knows that your proper sphere is vituperation. Your scolding
against me I regard as the vaporings of an old shrew. Your thunderous blows
do not intimidate me, but I dare tell you the truth. If you bark at me, I
will make you spit fire like a hell-hound. A perjured renegade ought to be
answered so that the whole world may know his name. For it is an old
saying that a wolf-roast demands canine sauces.

Hear me, Doctor Luther. I wish to propose for your consideration two
words which you apply to the princes who are opposed to you, namely, \textit{sacrileger}
and \textit{fool}. You, dishonorable and carnal wretch, fly at your opponents
as if they were sacrilegious despoilers of churches. But who was it that
sacrilegiously robbed the property of Christ, which emperors, kings, princes,
noblemen, citizens, and peasants, inspired by ardent Christian charity and
cherishing His sacred Passion, donated of old to the monasteries, parish-churches,
and altars? Verily, tell me truly, Squire Martin, Doctor Luther,
who despoiled the poor village parsons of their wretched income? Tell me,
Doctor Swinetrough (\textit{Säutrog}) Luther, who despoiled God of so many thousands
of souls during the last twelve years and sent them to Lucifer in hell?
Alas, that arch-murderer of souls, Luther! Who robbed Christ of His spouses,
the consecrated virgins, many of whom had served God for years in the
religious life, and forced them into a miserable, wretched, and erroneous life,
so that they now run about in great want and in disgrace? Fie on you,
Martin Luther, you perjured, sacrilegious whore-master of runaway monks
and nuns, apostate priests and all apostates! Who robbed and despoiled the
Roman Emperor, that beloved, innocent flower of Christianity, Charles V,
and the kings, princes, and lords, of honor and obedience on the part of their
subjects, by spreading false, seditious, and damnable writings and doctrines?
It was you, you execrable wretch! Who is responsible for the many thieves
and rogues that now infest almost every nook and corner of the land? It is
you, Doctor Luther!

Do you not know from Holy Writ what reward is in store for robbers?
Luther, your end and reward are known only to God.

The fruits and splendors of the new gospel are to be measured by other
standards. Who among you is able to check the growing drunkenness and
marital infidelity, the insubordination towards parents and masters? How
many violations of property, what crimes of highwaymen and prowling
thieves! This is the liberty which you have given to them: these are the
noble fruits of your teaching! You owe your success to the false liberty
which you have proclaimed.

You can well afford to preach as you do, for you enjoy the support of
your government. If you had not been favored by the Elector, much bloodshed,
revolt, and dissension, and the resurgence of all the old-time heresies
would have been avoided. Your place is the expected church council. There
you would meet men who are competent to dispute with you. But you prefer
to hurl your invectives from a safe place.

You say you are obliged to speak out because you are a doctor. Now,
I well know the oath which the doctors of Sacred Scripture are obliged to
take. It binds them to teach in conformity with the Church. You have
violated this oath. T fear you will be promoted in a school of which the devil
is rector; for, like you, he is all bluster and tumult.

You boast of your proficiency in German. If the German style of your
speeches is so subtle, then I know doctors in the country who are quite superior
to you in the vehemence of their German rhetoric, especially when
they are under the influence of strong drink. Ofttimes they beat with their
fists. If you were among them, you would not be taken for a doctor, but
for a swine-herd. You try by the raving fire of your words to start
a conflagration; but when the fire meets a solid rock, it merely covers it with
black soot, but cannot damage it.

You boast in your addresses that St. John Hus predicted your coming.
Hus is holy in the same manner in which he is holy who calls him a saint
or canonizes him, or has him prophesy about himself. Both he and you are
arch-heretics. It would have been better had you disputed with the learned
Master Erasmus of Rotterdam after he told you the
truth in his “Hyperaspistes.” He has so attuned his chords that you
could not reply. When you cease barking, you become silent, but you keep on sneaking about
treacherously like a mad dog.

Whilst the Peasants’ War was on, you delivered powerful speeches. How
many have lost their property and their lives, their bodies and their souls, as
a result of your false, seductive writings and your Satanic doctrines and
sermons! All the rebellious peasants who have been slain in that war will rise
up against you on Judgment Day and exclaim: “Woe to you!” You are an
arch-assassin of souls and bodies. You are guilty of their death, even as Pilate
is guilty of Christ’s. It is you who are a desperate, perjured blood-hound, not
they whom you call by this name to-day.

The whole world knows that your conscience upbraids you day and night,
and that you can nevermore be glad and enjoy peace of mind, even though
exteriorly you display a joyful disposition towards your Catherine von Bora.

In ten years you have been unable to settle in your own mind what you
will or ought to believe. You do not know year from now--your own writings
to-day what you will believe a prove it. You cannot come to an agreement
even with your own followers as to whether you intend to abide by
your present doctrines or invent others. As a consequence, the poor have
become to obfuscated that they scarcely know what to believe, and
almost every village has seen the rise of a new and distinct sect. What
prudent man, therefore, would adhere to you and desert the Catholic Church to
join your ranks? You have written that everybody ought to preach the
Gospel as he understands it. Now Karlstadt, Zwingli, Bucer, Capito,
Oecolampadius, the Anabaptists, Hubmaier, Hut, Müntzer, and many
others have followed your advice and preached according to their own judgment, thus
extending the misery which you have set afoot. You ought to restore order
instead of continuously heaping maledictions upon the papists.

Hear me, you arch-blasphemer; you admit before all the world that, when
in your daily prayer you say: “Hallowed be Thy Name!” you are perforce
compelled to exclaim: “Accursed, damned, reviled be the name of the papists
and all blasphemers.” Who blasphemes more outrageously than you? You
boast that you are a Christian doctor and preacher. Do you not know that
Holy Scripture teaches you not to curse your enemies, but to love them for
God’s sake? You should pray for them, as Christ did when He hung upon the
cross. You curse the papacy when you utter the words: “Thy kingdom
come!”, but do not forget that no one on earth is so much opposed to the
kingdom of Christ as you are. You would dethrone Him if it were possible.
Your followers have already cast His images out of the churches and demolished
them.

Your writings display the temerity and folly of a poor apostate who has
gone astray. If you want to see an arrogant fellow, behold, here is the most
arrogant of all. If you wish to see a destroyer of other people’s property,
behold, here is one from whom even the property of beggars is not safe. If you
want to see an unchaste man, behold, here is one who violates the chastity of
a nun. If you want to see an agitator and a rebel, behold, here is one guilty
of a hundred thousand murders and homicides, one who despises all spiritual
and secular authority and stands convicted as a falsifier of Holy Writ.

In you we find the source of every malice, evil, sin, and infamy; an unadulterated rogue.

Had not St. Paul written of Antichrist as he did, I should believe that you
were he; but undoubtedly you are his precursor. They are right who maintain
that Luther is certainly possessed by the devil. I believe that the whole
legion of devils which Christ drove out of the possessed and permitted to
enter into a herd of swine, is in you. As the devils made the swine so mad
that they drowned themselves in the sea: so the legion of devils has made
your monkish cranium so mad and giddy that you are unable to enjoy any
peace by day or night, until the Christians of all parties and places are become
confused and perplexed and finally plunged into the abyss of hell.

May you and your erring followers by the grace of God be brought back
to the way of truth! Would that you had a sincere desire to return! May
God Grant you repentance for your sins here below, and eternal happiness
to us all!

These are freely rendered extracts from the terrible excoriation
delivered against Luther from Dresden, especially by the sturdy
pastor, Arnoldi, in the course of the recriminations that followed the
diet of Augsburg. The language exceeds the bounds of propriety--
which is pardonable because of the heat of controversy--and illustrates
the fact that those who had become indignant at Luther freely
imitated the rhetorical tricks of their antagonist. At all events this
denunciation is a historical monument of an age when fierce contrasts
reached their climax. Although not by any means exempt from
passion, it reveals, by its very ardor, as no description could, the abhorrence
with which Catholic spokesmen viewed Luther, as he appeared to them in
his actions and writings, and the schism produced
by him in peaceful Germany.
